% Emacs settings: -*-mode: latex; TeX-master: "manual.tex"; -*-

\chapter{The component library}
\label{s:components}
This section is devoted to a description of components included in
the component library.  The component library is
maintained by the Ris\o{} group. All components were written at Ris\o\
except the chopper components in
sections~\ref{s:chopper} and~\ref{s:first_chopper} which have been
kindly contributed by Philipp Bernhardt, Lehrstuhl f\"ur
Kristallographie und Strukturphysik; and the Source\_Optimizer
(section~\ref{s:sourceoptimizer}), Monitor\_Optimizer
(section~\ref{s:monitoroptimizer}), and Monitor\_nD
(section~\ref{s:monitornd}) components which were written by Emmanuel
Farhi, Institute Laue-Langevin.
Users are encouraged to send
contributions to us for inclusion in future releases.

In the explanations of the individual components we will use
the usual symbols {\bf r} for the position $(x,y,z)$ of the particle
(unit m), and {\bf v} for 
the particle velocity $(v_x, v_y, v_z)$ (unit m/s).
Another frequently used symbol is 
the wave vector ${\bf k} = m_{\rm N} {\bf v}/\hbar$ , where
$m_{\rm N}$ is the neutron mass. {\bf k} is usually given in
\AA$^{-1}$, while neutron energies are given in meV.
In general, vectors are denoted by boldface symbols.
Subscripts "i" and "f" denotes "initial" and "final", respectively,
and are used in connection with the neutron state before and after
a scattering event.
Note that all mentioning of component geometry refer to
the local coordinate system of the individual component.

The source code for components is listed 
in Appendix~\ref{compcode}. 
The components follow the same numbering
in the Appendix as in the main text, \textit{e.g.} 
component \textbf{Arm}, subsection \ref{explain:arm},
appears in the Appendix as \ref{code:arm}.
Source code for many of the more trivial components are not included in
this manual. All sources may be found in the \verb+lib/mcstas/+
subdirectory of the McStas installation; the default is \verb+/usr/local/lib/mcstas/+.

\input{source}
\input{simple}
\input{guide}
\input{chopper}
\input{detector}
\input{monochr}
\input{powder}
\input{inelastic}
