% Emacs settings: -*-mode: latex; TeX-master: "manual.tex"; -*-

\chapter{New features in \MCS\ versions 1.7 - \version\ }
\label{c:changes}

Many new features have been implemented in \MCS\ version \version\  since
version 1.5 (no version 1.6. manual was written, the most important being that
\MCS\ can now compile and run in a Windows enviroment.  The list of changes
given here may be particularly  useful to experienced \MCS\ users who do not
wish to read the whole  manual, searching only for new features. A global
upward compatibility  exists for all changes, and thus all previous components
and  instruments should work as before. Changes are labelled as 1.7 or 1.8 to 
indicate in which \MCS\ version they appeared.

\section{General}
\label{s:new-features:general}
As of \MCS\ version \version\, the license covering the \MCS\ package
is the GNU General Pubclic License. See \verb+http://www.gnu.org+ for details.

\section{Kernel} 
\label{s:new-features:kernel}
\index{Kernel}

The following changes concern the 'Kernel' (i.e. the \MCS\ meta-language and program). See the dedicated chapter in the {\it User manual} for more details.

\begin{itemize}
\item \MCS\ can now compile for Windows!
\item Instrument parameters may now have default values, the same way as components do 
    (appeared in 1.8). \index{Parameters!Optional, default value}
\item An instrument source file may contain \texttt{EXTEND} \%\{ \ldots \}\% C
    blocks just after the usual \texttt{AT} \ldots \texttt{ROTATED} \ldots
    keywords, to extend the behaviour of existing components, without touching
    their code. All local component variables are  available. This may for
    instance be used to add a new 'color' to neutrons,  i.e. assign a new
    characteristic variable to the neutron (appeared in 1.7).
    \index{Keyword!EXTEND}
\item Component instances in an instrument file may be \texttt{GROUP}'ed into 
    exclusive assembly, i.e.\ only one component of the group will intercept
    the neutron ray, the rest will be skipped. This is useful for multi
    monochromators multi detectors, multiple collimators, etc.  After the
    \texttt{ROTATED} keyword, the keyword \texttt{GROUP} should be added
    followed by a group name (e.g. \texttt{GROUP} MyGroup)  (appeared in 1.7).
    \index{Keyword!GROUP}    
\item The instrument and components may have string (\verb+char*+) setting
    parameters. For components, their length is limited to 1024 characters
    (appeared in 1.7). \index{Parameters!Instruments} \index{Parameters!Setting}
\item In both components and instruments, the \texttt{FINALLY} section, that is
    executed at the end of simulations, has been supplemented with a new
    \texttt{SAVE} section. This latter is executed at simulation end (just
    before the \texttt{FINALLY} section), but also each time an intermediate
    save is required (e.g.\   when a 'kill -USR2 \$pid' is used under Unix, see
    section \ref{s:new-features:run-time})  (appeared in 1.7). 
    \index{Keyword!SAVE} \index{Keyword!FINALLY} \index{Signal handler!USR2 signal}
\item Components may have a \texttt{SHARE} section, which is imported only once
    per type of component. \texttt{SHARE} has the same role as
    \texttt{DECLARE}, but is useful when several instances of the same
    component is used in a single simulation  (appeared in 1.7).
    \index{Keyword!SHARE} \index{Keyword!DECLARE}
\item The component files may have some \texttt{\%include} inside '\%\{ \}\%'
    \texttt{DECLARE} or \texttt{SHARE} C blocks. The files to include are
    searched locally, and then in the library. If an extension is found, only
    the specified file is included, else both  .h and .c are embedded unless
    the --no-runtime has been specified. As in previous releases, the
    instrument files can embed external files, both in C blocks  and in the
    instrument parts (\texttt{DECLARE}, etc.) (appeared in 1.7).
    \index{Keyword!%include}
\item The \texttt{PREVIOUS} keyword, to be used after \texttt{RELATIVE} in
    place of component instance names refers to the preceeding component, and
    does not require to actually know  its name. Similarly the
    \texttt{PREVIOUS(n)} keyword refers to the $n$-th preceeding component
    (appeared in 1.8). \index{Keyword!PREVIOUS}
\end{itemize}

\section{Run-time} 
\label{s:new-features:run-time}
\index{Library!Run-time}

Some important modifications were done to the 'Run-time' library 
(i.e.\ the functions used in the instrument program). 
Some details may be found in {\it Running} chapter of the {\it User manual}
as well as in the appendix~\ref{c:kernelcalls}.

\begin{itemize}
\item A global gravitation handling is now available, by setting the \verb+-g+ flag.
\item Many output formats are available for data. Use the
    \verb+--format="format"+ flag, e.g. \verb+--format="Scilab"+.  The default
    format is McStas/PGPLOT, but may be specified globally using the
    \verb+MCSTAS_FORMAT+ environment variable.  See section~\ref{s:run-format}
    for details (appeared in 1.7). \index{Environment variable!MCSTAS\_FORMAT}
    \index{Data format}
\item It is possible to save 3D data arrays, by calling the DETECTOR\_OUT\_3D
    macro. (handled as 2D by \verb+mcplot+ by ignoring the 3rd dimension)
    (appeared in 1.7).  \index{Library!Run-time!DETECTOR\_OUT}
\item The C type of the 'number of events' array in monitors (usually named
    \verb+L_N+) was changed from \verb+int+ to \verb+double+, to avoid overflow.
    All 'home-made' monitors should be updated accordingly (appeared in 1.7).
\item The \verb+USR2+ signal generates an intermediate save for all monitors,
    during the simulation (executes the \texttt{SAVE} section). The \verb+USR1+
    signal still gives informations (appeared in 1.7). \index{Signal handler!USR1 signal}
\item New \verb+randvec_target_rect+ and \verb+randvec_target_rect_angular+
    functions now focus  on a rectangle (more efficient than the former
    \verb+randvec_target_sphere+) (appeared in 1.7 and 1.8).
    \index{Library!Run-time!randvec\_target\_rect}
\end{itemize}


\section{Components and Library} 
\label{s:new-features:components}
\index{Components} \index{Library!Components}
  
  We here list some of the new components (found in the \MCS\ \verb+lib+ directory) 
which are detailed in the {\it Component manual}, also mentioned in
the {\it Component Overview} of the {\it User Manual}.
  
\begin{itemize}
\item All components now have a valid header for \texttt{mcdoc} automatic documentation, as well as usage example.
\item \texttt{contrib} This directory now contains contributed
    components. Those that will be found to be highly stable will then
    go into the other component categories. Contributed components
    (\verb+Guide_honeycomb+, \verb+Guide_tapering+,
    \verb+Guide_curved+, \verb+ISIS_moderator+) have been
    placed there (appeared in 1.7 and 1.8). \index{Library!Components!contrib}
\item \texttt{data} This directory now contains example data files mainly for Laue diffraction, transmission and reflection files. \index{Library!Components!data}
\item \texttt{doc} The documentation is now included in the \verb+doc+ directory. \index{Library!Components!doc}
\item \texttt{example} A new \verb+example+ directory contains instrument examples, available from the \verb+mcgui+ 'Neutron site' menu Tool.\index{Library!Instruments}
\item \texttt{misc} Updated \verb+Vitess_input+ and \verb+Vitess_output+ components according to Vitess >= 2.3 Neutron structure.\index{Library!Components!misc}
\item \texttt{misc} A \verb+Progress_bar+ component may be positioned anywhere in the simulation and displays the simulation percentage. It may also save temporary data files regularly so that one can have a look \index{Tools!mcplot} (using \verb+mcplot+) at the results on-the-fly before the end of the simulation.
\item \texttt{monitors} The \verb+Monitor_nD+ component may be used as a replacement for all monitors. It may have automatic limits mode for either all or selected monitored variables. It may also plot banana monitors for \verb+mcdisplay+ and monitor something else than the intensity, e.g. the mean energy on a XY psd (appeared in 1.7). \index{Library!Components!monitors} A bug (atan2) was corrected in \verb+PSD_monitor_4PI+. \index{Library!Components!monitors}
\item \texttt{obsolete} Some components were renamed by categories for better sorting. Old components are still available, but not maintained anymore. We highly encourage to avoid using these when writing new instrument definitions. \index{Library!Components!obsolete}
\item \texttt{optics} components were renamed by categories, starting with
   Guide\_\ldots, Monochromator\_\ldots, Filter\_\ldots etc so that sorting is
   easier (appeared in 1.7). \index{Library!Components!optics}
\item \texttt{optics} Monochromator\_curved can read reflectivity and transmission tables (appeared in 1.7).   
\item \texttt{optics} Guide\_gravity can handle a 2D array of channels, and
    has options for subdivisions in length, chamfers and wavyness.
\item \texttt{optics} Filter\_gen can read a table from a file and affect the neutron
    beam (replaces the obsolete \verb+Flux_adapter+). It may act as a filter or a
    source (appeared in 1.7).    
\item \texttt{optics} Bugs were corrected in the following components: Bender, Chopper, Guide\_channeled. Some similar components were gathered. New components were added: Guide\_gravity, Guide\_wavy, Filter\_gen. \index{Library!Components!optics}
\item \texttt{samples} can now target towards any component, given its index 
    (no need to compute \verb+target_x|y|z+ vector, use e.g. \verb+target_index=1+). 
    Position an \verb+Arm+ at the focusing position when targetting to 
    centered components (appeared in 1.7). \index{Library!Components!samples}
\item \texttt{samples} Rectangular focusing has been implemented
  (instead of circular) in most components. More sample shapes are
  available. A bug has been corrected in Single\_crystal. A Powder2
  component (2 rings) has been added. Sans\_spheres is a new sample
  component for small angle scattering. Phonon\_simple is a new sample
  for phonon scattering. (appeared in 1.8). \index{Library!Components!samples}
\item \texttt{share} Many dedicated libraries are now available as shared code for reading tables,
    handling data files and monitors. These are C functions to be \texttt{\%include}d
    into components (see e.g. \verb+MCSTAS/monitors/Monitor_nD.comp+) (appeared in 1.7).
    \index{Library!Components!share}
\item \texttt{sources} The \verb+Source_gen+ and \verb+Source_Maxwell_3+ components may now simulate all kinds of continuous neutron sources. Some bugs were corrected in most sources in order to have an isotropic neutron emission. The \verb+Virtual_input+ and \verb+Virtual_output+ may load and save neutron event files (beware the size of the generated files !). Format may be text or binary (appeared in 1.7). \index{Library!Components!sources}
\end{itemize}

\section{Documentation}
\label{s:new-features:documentation}
As of \MCS\ \version\ an improved tutorial has been integrated
into to the graphical user interface of \MCS\. The content of the
tutorial is also available in appendix \ref{tutorial}.

\section{Example instruments}
\label{s:new-features:instruments}
As mentioned above, \MCS\ \version\ has various improvements regarding
distribution of instruments. Firstly, \verb+mcgui+ has a new special
menu for instruments distributed with \MCS and secondly instruments can
now have default values and thirdly a new \MCS\ tutorial has
integrated into the user interface. These three new features together
proves a solid starting point for new users of \MCS\. To further
increase the value of the package, we have decided to include selveral
example instruments contributed by \MCS\ users. The full list of 
example instruments are
\begin{itemize}
\item{\verb+vanadium_example+, improved version used in the \MCS\
    tutorial.}
\item{\verb+linup-1+-\verb+linup-7+, legacy instruments of the Ris\o\
    TAS1 instrument from a historical release of \MCS.}
\item{\verb+h8_test+, model of the Brookhaven H8 triple-axis, written
    by McStas team member Emmanuel Farhi}
\item{\verb+ILL_D9+, a model of the ILL D9 hot four-circle
    diffractometer, contributed by Chris Ling}
\item{\verb+Hetfull+, a model of the ISIS HET spectrometer,
    contributed by Dickon Champion}
\item{\verb+focus_psi+, a model of the PSI FOCUS spectrometer,
    contributed by Uwe Filges}
\item{\verb+prisma2+, a model of the ISIS prisma2 spectrometer,
    written by Kristian Nielsen and Mark Hagen from a historical release of \MCS.}
\item{Test instruments \verb+ISIStest+ for testing the new
    \verb+ISIS_moderator+ component, \verb+SANS+ for testing the new
    \verb+Sans_spheres+ component and \verb+Test_Phonon+ for testing
    the new \verb+Phonon_simple+ component, all contributed by the
    corresponding component authors.} 
\end{itemize}



\section{Tools, installation}
\label{s:new-features:tools}
\index{Tools}
\index{Installing}
  A renewal of most \MCS\ Tools, used to edit and display instrument or
  results,  has been undertaken, aiming at proposing alternatives to the
  Perl+PerlTk+PGPLOT+PDL  libraries.
  
  Quite a lot of work was achieved in order to solve the installation problems
   that have been encoutered so far. A fully working \MCS\ distribution now
   only requires a C compiler, perl, perl-Tk and one of Matlab, Scilab and
   (PGPLOT, perl-DL). The Plotlib Scilab library has  been included in the
   package, and does not need to be installed separately anymore.
  
  This has improved significantly the portability of \MCS\ and thus simplified
  the installation of the package. Details about the installation and the
  available tools are given in the {\it Installing \MCS} chapter of the {\it User Manual}.

\begin{itemize}
\item The list of required packages for a complete \MCS\ installation is now a C
    compiler, Perl, PerlTk and Scilab or Matlab.
\item Matlab, Scilab and IDL may read directly McStas results if the simulation
    was executed with the \verb+--format="..."+ option 
    (see~\ref{s:new-features:run-time} changes). The former PGPLOT interface is
    still supported (appeared in 1.7).  \index{Tools!Matlab}
    \index{Tools!Scilab}  \index{Tools!IDL} \index{Tools!Perl libraries}   
\item\verb+mcgui+ can now perform \verb+mcrun+-like parameter scans
    directly from the gui (appeared in 1.8).
\item\verb+mcgui+ has now a 'Neutron site' menu which enables to load directly one of the
    instrument examples (appeared in 1.8).
\item \verb+mcrun+ can generate scan results in all formats (appeared in 1.8). 
   \index{Tools!mcrun}
\item \verb+mcrun+ has a \MCS\ self test procedure available (appeared in 1.8). 
   \index{Tools!mcrun}
\item \verb+mcplot+, \verb+mcdisplay+, \verb+mcgui+ are now less dependent on the
    \verb+perl/PDL/pgplot+ installed versions and fully work with Matlab/Scilab (appeared in 1.7). \index{Tools!mcplot}
\item \verb+mcplot+ can plot a single simulation data file.
\item \verb+mcplot+, \verb+mcresplot+, \verb+mcdisplay+ can output GIF, PS and color
    PS files. They also have integrated help (-h options), and may generate output
    files in a non interactive mode (read data, create output file, exit) (appeared in 1.7).
\item \verb+mcplot+ and \verb+mcdisplay+ work with Matlab, PGPLOT and Scilab plotters
   (depends on the \verb+MCSTAS_FORMAT+ environment variable, or -pPLOTTER option, or
   PGPLOT if not set) (appeared in 1.7).
\item \verb+mcplot+ may display parameter scan step results in all formats
   (appeared in 1.8).  \index{Tools!mcplot}
\item \verb+mcstas2vitess+ enables to convert a McStas component into a
   Vitess~\cite{vitess_webpage} one (appeared in 1.7). \index{Tools!mcstas2vitess}
\item \verb+mcresplot+ can plot projections of the 4D resolution function of an 
   instrument obtained from the \verb+Res_sample+ and \verb+Res_monitor+ components.
   In version \version, it only works with the McStas/PGPLOT format, 
   but a port for Scilab/Matlab is under way (appeared in 1.7). \index{Tools!mcresplot}
\item \verb+mcdoc+ can now display the pdf version of the manual, an
  HTML catalog of the current library, as well as help for single
  components. \index{Tools!mcdoc} The \verb+mcdoc+ functions have been
  closely integrated into \verb+mcgui+ (appeared in 1.7). 
\item \verb+mcdoc+ can document automatically instruments in the same way as components
 (appeared in 1.8).
\item \verb+mcconvert+ is a new tool to convert \MCS\ result text files from Matlab to Scilab, 
  and from Scilab to Matlab formats (appeared in 1.8). \index{Tools!mcconvert}
\end{itemize}

\section{Future extensions}
\label{s:future}
The following features are planned for the oncoming releases of \MCS :
\begin{itemize}
\item Support for Matlab and Scilab in \verb+mcresplot+.
\item Support for MPI (parallel processing library) in the runtime
\item Language extension 'JUMP' for enabeling loops, 'teleporting'
  etc. in instrument descriptions.
\item Concentric components.
\item Polarised components and magnetic field computation components.
\item Optimize a set of parameters for a better flux and/or resolution on a given monitor.
\end{itemize}








