\chapter{Samples}
\index{Library!Components!samples}

This class of components models the sample of the experiment.
This is by far the most challenging part of a neutron scattering
instrument to model. However, for purpose of simulating
instrument performance, details of the samples are rather unimportant,
allowing for simple approximations. On the contrary, for full
virtual experiments it is of importance to have realistic and
detailed sample descriptions. \MCS\ contains both simple and detailed
samples, although there is still room for much more development.

We first consider incoherent scattering. The simple component {\bf Vanadium}
performs both incoherent scattering and absorption.

An important component class is elastic Bragg scattering from an ideal powder.
The component {\bf PowderN} models a powder scatterer with reflections 
given in a LAZY/Crystallographica file.
The component includes absorption and incoherent scattering. 

Next type is Bragg scattering from single crystals.
The simplest single crystals are in fact the monochromator components
{\bf Monochromator\_flat} and {\bf Monochromator\_curved},
which were presented in the section on optics.
The monochromators are models of a thin mosaic crystal
with a single scattering vector perpendicular to the surface.
Much more advanced, the component {\bf Single\_crystal}
is a general single crystal sample (with multiple scattering) that allows
the input of an arbitrary unit cell and a list of structure factors, read
from a LAZY/Crystallographica file.
This component also allows anisotropic mosaicity 
and $\Delta d/d$ lattice space variation.

Isotropoic small-angle scattering is simulated in {\bf Sans\_Spheres},
which models scattering from a collection of hard spheres. A more general
SANS sample is under development. 
In addition, a reflectometry sample will soon be developed.

Inelastic scattering from a dispersion is exemplified by
the component {\bf Phonon\_simple}, which models a single phonon dispersion.

For a more general sample model, the {\bf Isotropic\_Sqw} component 
is able to simulate all kinds of isotropic materials: 
Liquids, glasses, polymers, powders, etc, with $S(q,\omega)$ table
specified by an input file.
Physical processes include coherent/incoherent scattering, 
both elastic and inelastic, with absorption and multiple scattering. 
Moreover, this component may be used concentrically, 
to model a sample environment. 
Thus it may handle most samples except single crystals.

In general, all samples are assumed to be homogeneous. There would also be
potential in developing an inhomogeneous sample, e.g. with
spatially varying lattice constant, relevant for stress/strain scanners.
Inhomogeneously absorbing sample for tomography could also be possible.
Further, no polarization effects are yet taken into account in any
of the samples.

\begin{table}
  \begin{center}
  {\let\my=\\
    \begin{tabular}{|c|cc|cc|c|c|}
    \hline
    Sample        & \multicolumn{2}{c|}{Coherent} & \multicolumn{2}{c|}{Incoherent} &&\\
    Process       & Elastic & Inelastic & Elastic & Inelastic & Absorption & Multi. Scatt.\\
    \hline
    Phonon\_simple&         & X         &         &           & X & \\
    Isotropic\_Sqw&  X      & X         & X       & X         & X & X \\
    PowderN       &  N lines&           &         &           & X & \\
    Sans\_spheres &  colloid&           &         &           & X & \\
    Single\_crystal& X      &           & X       &           & X & X \\
    V\_sample     &         &           & X       &           & X & \\
    \hline
    \end{tabular}
    \caption{Processes implemented in sample components}
    \label{t:sample-process}
  }
  \end{center}
\end{table}

\input{vanadium.tex}         \newpage
\input{powder2.tex}          \newpage
\input{Single_crystal.tex}   \newpage
\input{sans.tex}             \newpage
\input{phonon.tex}           \newpage
%\input{LSCO.tex}            \newpage
\input{isotropic_sqw.tex}    

