% Emacs settings: -*-mode: latex; TeX-master: "manual.tex"; -*-

\chapter{Source components}
\label{c:source}
\index{Sources}
The main function of the source components is to determine a set of initial
parameters $({\bf r}, {\bf v})$, or equivalent (${\bf r}, v, \Ombold $),
for each neutron. This is done by Monte Carlo choices. 
In the current sources no polarization dependence is implemented, 
whence we let ${\bf s}=(0,0)$.

The sources to be presented in the following all make their Monte Carlo
choices for initial position and direction on the basis of a 
uniform distribution, with the exception of Souce\_Div, where the 
distribution of initial divergence (deviation between the nominal and
actual direction) is Gaussian. The shape of the source may be a disk (Moderator,  Source\_adapt, Source\_flat and Source\_flux), a rectangle (Source\_div), a cube (ESS sources, Source\_Maxwell\_3)or even a box/cylinder (Source\_gen). The choice of energy distribution
varies from a flat distribution in Source\_flat, to the more realistic
weighted sum of three Maxwellians in Source\_Maxwell\_3 and Source\_gen.
As neutron energy is in principle unlimited, all sources require
energy/wavelength windows to be specified.

For time-of-flight sources, the initial neutron time, $t$ (in s), is
an essential parameter. The choice of this is being made on basis of
detailed analytical expressions, see the respective components. 
For other sources, the initial neutron time is set to zero.

The flux of the sources deserves special attention. The total flux is defined
as the sum of weights of all emitted neutron rays during one simulation 
(the unit of weight is thus neutrons per second), if all neutron energies
were allowed. In most sources, the total flux is a required parameter.
For the simple sources Source\_flat and Source\_flat\_lambda (mostly used
for test purposes), the flux may be omitted. In this case the total integrated flux is set to unity.

\section{Source\_flat: A circular continuous source with a flat energy spectrum}
\label{source-flat}
\index{Sources!Source\_flat}
\component{Source\_flat}{System}{ $r_{\rm s}$, $z_{\rm f}$, $w$, $h$, $E_0$, $\Delta E$}{None}

This component is 
a simple continuous source with a energy distribution which is uniform
in the range $E_0 - \Delta E$ to $E_0 + \Delta E$.
The component is not used for time-of-flight simulations,
so we put $t=0$ for all neutron rays.

The initial neutron ray position is chosen randomly from within a
circle of radius $r_{\rm s}$ in the $z=0$ plane. 
This geometry is a fair approximation
of a cylindrical cold/thermal source with the beam going out along
the cylinder axis.

The initial neutron ray direction is focused within
a solid angle, defined by a rectangular target of width
$w$, height $h$, parallel to 
the $xy$ plane placed at $(0,0,z_{\rm f})$. 

The total weight of the neutron beam is set to the
solid angle of the focusing opening divided by $4\pi$.

\section{Source\_flux: A circular continuous source with a flat energy spectrum and absolute flux}
\label{source-flux}
\index{Sources!Source\_flux}

\section{Source\_flat\_lambda: A continuous source with flat
  wavelength spectrum}
\label{source-flat-lambda}
\index{Sources!Source\_flat\_lambda}

\component{Source\_flat\_lambda}{System}{ \textit{radius, dist, xw, yh, lambda\_0, d\_lambda}}{None}

The component {\bf Source\_flat\_lambda} is similar to the Source\_flat
component, except that the spectrum is flat in wavelength, rather
than in energy.

The input parameters for Source\_flat\_lambda are \textit{radius} to set
the source radius; \textit{dist}, \textit{xw}, and \textit{yh}
to set the focusing as for Source\_flat; and \textit{lambda\_0} and
\textit{d\_lambda} to set the range of wavelength emitted (the range
will be from $\textit{lambda\_0} - \textit{d\_lambda}$ to
$\textit{lambda\_0} + \textit{d\_lambda}$).

IN GENERAL: HOW DO WE HANDLE MATH SYMBOLS IN THE COMPONENT DESCIPTIONS? THE 
FIRST TWO COMPONENT DESCRIPTIONS ARE WIDELY DIFFERENT!

\section{Source\_flux\_lambda: A continuous source with a flat wavelength spectrum and absolute flux}
\label{Source_flux_lambda}

The component {\bf Source\_flux\_lambda} is a variation on the
Source\_flat\_lambda component. The only difference is the possibility
to specify the absolute flux of the source. The specified flux is used
to adjust the initial neutron weight so that the intensity in the
detectors is directly comparable to a measurement of one second on a
real source with the same flux. This also makes the simulated detector
intensities independent of the number of neutron histories simulated,
easing the comparison between different simulation runs (though of
course more neutron histories will give better statistics).

The flux~$\Phi$ is the number of neutrons emitted per second from a
one~cm$^2$ area on the source surface, with direction within a a one
steradian solid angle, and with wavelength within a one {\AA}ngstr{\o}m
interval. The total number of neutrons emitted towards a given diaframe
in one second is therefore
$$ N_{\rm total} = \Phi A \Omega \Delta\lambda $$
where $A$ is the source area, $\Omega$ is the solid angle of the
diaframe as seen from the source surface, and $\Delta\lambda$ is the
width of the wavelength interval in which neutrons are emitted (assuming
a uniform wavelength spectrum). If $N_{\rm sim}$ denotes the number of
neutron histories to simulate, the initial neutron weight $p_0$ must be set to
$$ p_0 = \frac{N_{\rm total}}{N_{\rm sim}} = 
    \frac{\Phi}{N_{\rm sim}} A \Omega \Delta\lambda $$

The input parameters for Source\_flux\_lambda are \textit{radius} to set
the source radius in meters; \textit{dist}, \textit{xw}, and \textit{yh}
to set the focusing as for Source\_flat; \textit{lambda\_0} and
\textit{d\_lambda} to set the range of wavelength emitted (the range
will be from $\textit{lambda\_0} - \textit{d\_lambda}$ to
$\textit{lambda\_0} + \textit{d\_lambda}$); and \textit{flux} to set the
source flux in units of ${\rm cm}^{-2} {\rm st}^{-1} \textit{\AA}$.


\section{Source\_div: A divergent source}

{\bf Source\_div} is a rectangular source which emits a
beam of a certain divergence around the main exit direction
(the direction of the $z$ axis).
The beam intensity and divergence are uniform over
the whole of the source, and the energy distribution
of the beam is uniform.

This component may be used as a simple model of the
beam profile at the end of a guide or at the sample
position.

The input parameters for Source\_div are the source dimensions
$w$ and $h$ (in m), the divergencies $\delta_h$ and $\delta_v$ (FWHM in degrees), 
and the mean energy $E_0$ and the energy spread $dE$ (both in meV).
The neutron energy range is $(E_0-dE; E_0+dE)$. 


\section{Moderator: A time-of-flight source}
The simple time-of-flight source component {\bf Moderator} resembles
the source component {\bf Source\_flat} described in \ref{sourceaim}.
Like {\bf Source\_flat}, {\bf Moderator} is circular and focuses
on a rectangular target. Further, the initial velocity is chosen
with a linear distribution within an interval, defined by the
minimum and maximum energies, $E_0$ and $E_1$,
respectively.

The initial time of the neutron is determined on basis of a 
simple heuristical model for the time dependence of the 
neutron intensity from a time-of-flight source.
For all neutron energies, the flux decay is assumed to be exponential,
\begin{equation}
\Psi(E,t) = \exp(-t/\tau(E)) ,
\end{equation}
where the decay constant is given by
\begin{equation}
\tau(E) = \left\{ 
\begin{array}{cc}
 \tau_0                               & ; E<E_c \\
 \tau_0 / [ 1 + (E-E_c)^2/\gamma^2 ]  & ; E \geq E_c
\end{array}
\right.
\end{equation}

The input parameters for {\bf Moderator} are the source radius, $r_{\rm s}$,
the minimum and maximum energies, $E_0$ and $E_1$ (in meV),
the distance to the target, $z_{\rm f}$, the dimensions of the target,
$w$ and $h$, and the decay parameters 
$\tau_0$ (in $\mu$s), $E_c$, and $\gamma$ (both in meV).


\input{Source_adapt}

\input{Source_Optimizer}
