% Emacs settings: -*-mode: latex; TeX-master: "manual.tex"; -*-

\addcontentsline{toc}{chapter}{\protect\numberline{}{Preface and acknowledgements}}
\chapter*{Preface and acknowledgements}
This document contains information on the Monte Carlo neutron
ray-tracing program \MCS\ version \version, an update to the initial
release in October 1998 of version 1.0 as presented in Ref.~\cite{nn_10_20}. The reader of this
document is supposed to have some knowledge of neutron scattering,
whereas only little knowledge about simulation techniques is
required. In a few places, we also assume familiarity with the
use of C, UNIX and of the world wide web (WWW).

It is a pleasure to thank Prof.~Kurt N.~Clausen for his continuous
support to this project and for having initiated the work in the first
place. 
%Both he and our other collaborators, Henrik M.\ R\o nnow and Mark
%Hagen have made major contributions to the project.  Also the
%contributions from our test users, the students Asger Abrahamsen, Niels
%Bech Christensen, and Erik Lauridsen, are gratefully acknowledged; they
%gave us an excellent opportunity to pinpoint a vast amount of serious
%errors in the test version.  Useful comments to this document itself
%have been given by Bella Lake and Alan Tennant.  
We have also benefited
from discussions with many other people in the neutron scattering
community, too numerous to mention here.

%Philipp Bernhardt contributed the two chopper components in
%sections~\ref{s:chopper} and~\ref{s:first_chopper}. Emmanuel Farhi
%contributed the components in sections~\ref{s:sourceoptimizer},
%\ref{s:monitornd}, and~\ref{s:monitoroptimizer}. We encourage other
%users to contribute components with manual entries for inclusion in
%future versions of \MCS.

This project has been supported by the European Union
through the XENNI program and the RTD ``Cool Neutrons'' and ``SCANS'' programs.

In case of any errors, questions, suggestions, %or other need for support should arise,
%do not hesitate to 
contact the authors at \verb+mcstas@risoe.dk+
or consult the \MCS\ WWW home page~\cite{mcstas_webpage}.

New features in \MCS\ version \version\ as compared to version 1.4 include

\begin{itemize}
\item A three-component representaton, (s$_x$, s$_y$, s$_z$), of the neutron spin has been added, 
which includes the possibility for simulations of polarisation instruments.
\item A signal handling system has been added, which gives the possibility for interacting with a
running simulation
\item The list of available kernel calls have been extended. They are normally used by users coding
new components. The new kernel calls include the name, position and orientation of components.
\item The \MCS\ manual is separated into two documents: this manual and a new manual containing
component descriptions. The first \MCS\ component manual will be released in October/November 2001.
\item The release contains many new components and updates of existing components. They will be 
documented in the new \MCS\ component manual, but information on all \MCS\ components may be found 
at the McStas web-page~\cite{mcstas_webpage}.
\end{itemize} 







