% Emacs settings: -*-mode: latex; TeX-master: "manual.tex"; -*-

\addcontentsline{toc}{chapter}{\protect\numberline{}{Preface and acknowledgements}}
\chapter*{Preface and acknowledgements}
This document contains information on the neutron scattering components 
which are the building blocks for describing instruments 
in the Monte Carlo neutron
ray-tracing program \MCS\ version \version . This is an update to the initial
release in October 1998 of version 1.0 as presented in Ref.~\cite{nn_10_20}. 
The reader of this
document is not supposed to have specific knowledge of neutron scattering,
but some basic understanding of the underlying physics is helpful in
understanding the theoretical background for the component functionality. 
For details about simulation techniques, we refer to 
the McStas system manual \cite{mcstasmanual}.
We assume familiarity with the use of 
the C programming language.

We especially like to thank Kristian Nielsen for laying a solid foundation
for the \MCS\ system, which the authors of this manual are using daily.
It is also a pleasure to thank Prof.~Kurt N.~Clausen for his continuous
support to \MCS\ and for having initiated the project.
%Both he and our other collaborators, Henrik M.\ R\o nnow and Mark
%Hagen have made major contributions to the project.  Also the
%contributions from our test users, the students Asger Abrahamsen, Niels
%Bech Christensen, and Erik Lauridsen, are gratefully acknowledged; they
%gave us an excellent opportunity to pinpoint a vast amount of serious
%errors in the test version.  Useful comments to this document itself
%have been given by Bella Lake and Alan Tennant.  
We have also benefited
from discussions with many other people in the neutron scattering
community, too numerous to mention here.

%Philipp Bernhardt contributed the two chopper components in
%sections~\ref{s:chopper} and~\ref{s:first_chopper}. Emmanuel Farhi
%contributed the components in sections~\ref{s:sourceoptimizer},
%\ref{s:monitornd}, and~\ref{s:monitoroptimizer}. 
The users who contributed components to this manual are acknowledged
as authors of the individual components. We encourage other
users to contribute components with manual entries for inclusion in
future versions of \MCS.

In case of any errors, questions, suggestions, 
%or other need for support should arise,
do not hesitate to 
contact the authors at \verb+mcstas@risoe.dk+
or consult the \MCS\ WWW home page~\cite{mcstas_webpage}.

Important developments on the component side in \MCS\ version \version\ 
as compared to version 1.4 (the last version of the component manual) include

\begin{itemize}
\item I do not know what to say...
\end{itemize} 

The \MCS\ project has been supported by the European Union, initially
through the XENNI program and the RTD ``Cool Neutrons'' program.
Over the last 3 years, \MCS\ was supported strongly through the
``SCANS'' program.






